\chapter{Hotmoka blockchains}\label{ch:hotmoka_blockchains}

The experiments of the previous chapters have been performed on a
Hotmoka node already existing online, part of a blockchain installed by somebody else.
Namely, we have used the Hotmoka node at \serverMokamint{}, that is part
of a blockchain based on a proof of space consensus algorithm~\cite{Spoto25}\index{proof of space},
and the Hotmoka node at \serverTendermint{}, that is part of
another blockchain, this time based on a proof of stake consensus algorithm~\cite{Kwon14}\index{proof of stake}.
Both are able to execute Hotmoka transactions, store smart contracts and interact with them,
but they are different for the way they create new blocks and
synchronize their state among all the nodes
of the blockchain they belong to. Namely, in the case of proof of space, each node
of the blockchain increases the likelihood of winning the right to create a new block
if it allocates a larger data structure on disk (typically, on SSD) than the other nodes of the
same blockchain. Instead, in the case of proof of stake, only a restricted, modifiable set of nodes
(called \emph{validators}\index{validator}) has the right to create new blocks,
in a round-robin way, and the prize earned for the creation
of a new block is proportional to the shares of validation power\index{validation power}
owned by each validator. Other nodes can just verify the transactions.
We do not enter into the (sometime religious) discussion about which consensus algorithm
is the best. Probably none of them, as each of them (and others, that we do not even speak about here)
has its advantages and drawbacks. What is important to highlight is that Hotmoka is not
bound to a specific consensus algorithm, but is rather a software layer for the execution
of smart contract transactions, that can be plugged on top of different consensus engines,
giving rise to distinct blockchains with distinct consensus algorithms.

This chapter shows how to build new blockchains or connect to existing blockchains,
both for proof of space and for proof of stake consensus. It will present this by using docker,
so that you do not need to install anything on your machine beyond docker itself.
Namely, there exist preconfigured docker\index{docker} images that allow one to run Hotmoka nodes of
a blockchain without preliminary, complex configuration. They are meant to simplify the
life of users who just want to install nodes with a standard procedure that should cover
most typical use cases. For less standard use cases, such docker images can be partially configured
or Hotmoka nodes can be installed with Java code, as it will be shown later in Ch.~\ref{ch:hotmoka_nodes}.

\section{Tokenomics}\label{sec:tokenomics}\index{tokenomics}

The way coins are minted and transaction fees are computed is the same in all implementations
of Hotmoka, independently from the underlying consensus algorithm. In particular,
there is an amount of coins $\mathit{minted}_h$ that gets minted for the creation of the
block $b_h$ at height $h$. This minted coins, plus the
$\mathit{fee}_h$ resulting from the execution of the transaction requests inside $b_h$,
go to the \emph{creators} of $b_h$. Who are these creators depends from the
consensus engine. For instance, for proof of space based on Mokamint, the creators are
a peer and a miner and $\mathit{minted}_h+\mathit{fee}_h$ gets split, equally, between the two.
For proof of stake based on Tendermint, instead, there is a single validator peer that creates
$b_h$ and gets all $\mathit{minted}_h+\mathit{fee}_h$ coins. Note that $\mathit{minted}_h$
are minted coins, while $\mathit{fee}_h$ are coins paid by who required to execute the
transactions: they are not minted.

The computation of $\mathit{minted}_h$ follows a decreasing hyperbolic curve. Namely, there
are consensus parameters, decided at the time of blockchain launch, that determine
$\mathit{minted}_h$:
%
\begin{itemize}
\item $\mathit{initial\_supply}$\index{initial supply} is the amount of coins that already exist at launch time.
Initially, they are all inside an account called gamete, typically
controlled by the person who launched the blockchain; this amount can well be zero;
\item $H$ is the \emph{height at final supply}, that is, the block height when all coins
will have been minted: coins will only be minted for blocks $b_h$ at height $0\le h<H$:
subsequent blocks will only distribute $\mathit{fee}_h$;
\item $\mathit{final\_supply}$\index{final supply} is the amount of coins that will exist, in total, at height $H$;
it must be $\mathit{final\_supply}\ge\emph{initial\_supply}$.
\end{itemize}
%
For instance, for the main network \texttt{\serverMokamintMainChainId{}} of Hotmoka (Fig.~\ref{fig:endpoints}),
the above parameters are $\mathit{initial\_supply}=0$, $\mathit{final\_supply}=10^{30}$
and $H=42048000$, as it is possible to verify with the command
\texttt{moka nodes manifest show --uri \serverMokamintMain{}}.

The hyperbolic curve that dictates how many coins get minted at block $0\le h<H$ is the following:
\[
  \mathit{minted}_h=\frac{6(\mathit{final\_supply}-\mathit{initial\_supply})(h^2-2hH+H^2)}{(H+1)(2H+1)H}.
\]
For instance, Fig.~\ref{fig:maresia_minting} reports how many coins \texttt{\serverMokamintMainChainId{}}
will mint at different block heights.
%
\begin{figure}[th]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
$h$ & when (days after launch) & $\mathit{minted}_h$ \\\hline\hline
0 & $0$ & 71347029418270905255751\\\hline
4320 & $1$ & 71332369822860396367870\\\hline
129600 & $30$ & 70907896753038047445079\\\hline
1576800 & $365$ & 66096333972020030822086\\\hline
3153600 & $2\times 365$ & 61046302046008043309452 \\\hline
4730400 & $3\times 365$ & 56196933640234942717850 \\\hline
6307200 & $4\times 365$ & 51548228754700729047280 \\\hline
7884000 & $5\times 365$ & 47100187389405402297742 \\\hline
9460800 & $6\times 365$ & 42852809544348962469235 \\\hline
11037600 & $7\times 365$ & 38806095219531409561761 \\\hline
40996800 & $26\times 365$ & 44591893386419315784 \\\hline
$\ge 42048000$ & $\ge(26\times 365 + 244)$ & 0 \\\hline
\end{tabular}
\end{center}
\caption{The amount of coins $\mathit{minted}_h$ that the \texttt{\serverMokamintMainChainId{}}
main network of Hotmoka will mint at different block heights $h$.
For the days after launch when a block will be minted, consider that \texttt{\serverMokamintMainChainId{}} is
programmed to create a block every $20$ seconds, on average.}
\label{fig:maresia_minting}
\end{figure}

\section{Hotmoka blockchains based on Mokamint (proof of space)}\label{sec:hotmoka_mokamint}

Hotmoka can run on top of a generic engine for proof of space consensus, called Mokamint~\cite{mokamint_github}.
The latter implements the networking and consensus layers of a blockchain, leaving the application
layer generic, so that it can be plugged on top of it. Hotmoka is one such application layer,
but others may exist as well. Mokamint implements a consensus algorithm based on proof of space,
derived from that of Burstcoin/Signum~\cite{Spoto25}. In such algorithm, the creation of a new
block required to identification of a data structure, called \emph{nonce}, from a set of
nonces precomputed on disk and held inside a big file called \emph{plot},
in order to minimize the waiting time for the creation of the next block.
The more nonces are precomputed on disk, the higher the likelihood of winning the competition for
the creation of the next blocks. Nonces could in principle be recomputed on the fly at each
block creation, but their computation is so complex that it becomes impossible to perform it
in the short time between the creaton of successive blocks. From this point of view, proof of space
is similar to Bitcoin's proof of work~\cite{Nakamoto08}, but the resource used for the competition
is different: SSD instead of CPU. Consequently, proof of space has a minimal energy cost and can be
run a cheap hardware as well. In particular, you can mine for Mokamint-based blockchains with a desktop machine
or even with a mobile phone or tablet (see the Mokaminter Android app~\cite{mokaminter_google_play}).
This becomes possible for Hotmoka as well, if run on top of Mokamint, as shown in this section.

\begin{figure}[t]
  \begin{center}
    \begin{tikzpicture}
      \draw [fill=verylightviolet,thick] (-0.5,-1.5) rectangle +(4,3);
      \draw (1.5,1.1) node {Mokamint};
      \draw (1.5,0.6) node {{\scriptsize (networking and consensus)}};
      \draw[-, thick] (1.5,-1.5) -- (1.5,-2.7);
      \draw (1.5,-2.7) node[scale=0.7] {\db};
      \draw (1.5,-3.6) node {database of blocks};
      \draw[<->, thick] (3.5,-1) -- (5.5,-1);
      \draw (7.8,-1) node {other blockchain peers};
      \draw (4.5,-1.2) node {{\scriptsize whispering}};

      \draw [fill=verylightyellow,thick] (0.5,-0.7) rectangle +(2,0.5);
      \draw (1.5,-0.45) node {mempool};
      \draw[->, thick] (-1.5,-0.45) -- (0.5,-0.45);
      \draw (-3.5,-0.45) node {transaction requests};

      \draw [fill=verylightred,thick] (7,0.35) rectangle +(2,0.8);
      \draw (8.0,0.75) node {miner(s)};

      \draw [fill=verylightgreen,thick] (0.3,3) rectangle +(2.4,0.8);
      \draw (1.5,3.4) node {Hotmoka};
      \draw[-, thick] (0.3,3.4) -- (-1,3.4);
      \draw (-1,3.4) node[scale=0.7] {\db};
      \draw (-1,2.5) node {database of states};

      \draw[->, thick] (3.5,1.0) -- (7,1.0);
      \draw (5.25,1.2) node {{\small challenges}};

      \draw[<-, thick] (3.5,0.5) -- (7,0.5);
      \draw (5.25,0.25) node {{\small answers}};

      \draw[->, thick] (1.2,1.5) -- (1.2,3);
      \draw[->, thick] (1.8,3) -- (1.8,1.5);
      \draw (1.5,2.25) node {{\scriptsize API}};
    \end{tikzpicture}
  \end{center}
  \caption{The architecture of a Hotmoka node based on the Mokamint proof of space engine.
    Mokamint peer, miner(s) and Hotmoka application are separate software
    components, possibly running in distinct machines connected through a network.}\label{fig:architecture}
\end{figure}

Fig.~\ref{fig:architecture} shows the architecture of a Hotmoka node running on top of Mokamint.
As you can see, Mokamint is taking care of networking and consensus and keeps the actual blockchain
inside its own database. It takes part in mining new blocks through one or more miners.
Miners are not necessarily \emph{local}\index{miner!local}, that is, providing themselves support for mining. They can also
be \emph{remote} miners\index{miner!remote}, that can be imagined as plugs to which external miners can connect to help
the Mokamint node create new node, in exchange of a part of the earned coins.
Moreover, each Mokamint node is a peer connected to other Mokamint peers, from which it can receive blocks through
whispering. Users can send transaction requests to a Mokamint node: initially, they will be stored
in a temporary area called mempool, from where they get eventually selected for inclusion in a block.
Hotmoka is a separate component that runs transactions inside its database of states. This database
is where Java objects, including accounts, are stored.

\begin{figure}[th]
  \begin{center}
    \begin{tikzpicture}[scale=1.1]
      \draw [fill=lightviolet,thick] (0,1) rectangle +(3,1.5);
      \draw[<->, thick,dashed] (1.5,0.25) -- (1.5,1);
      \draw [fill=yellow,thick] (0,-0.75) rectangle +(3,1);
      \draw (1.5,-0.25) node {{\scriptsize peer1 (Hotmoka)}};
      \draw (1.5,2) node {{\scriptsize peer1 (Mokamint)}};
      \draw (1.5,1.5) node {{\scriptsize 8dE4pHgtvFAXgB\ldots}};
      \draw [fill=lightviolet,thick] (0,4) rectangle +(3,1.5);
      \draw[<->, thick,dashed] (1.5,5.5) -- (1.5,6.25);
      \draw [fill=yellow,thick] (0,6.25) rectangle +(3,1);
      \draw (1.5,6.75) node {{\scriptsize peer2 (Hotmoka)}};
      \draw (1.5,5) node {{\scriptsize peer2 (Mokamint)}};

      \draw (4.3,6.75) node {\leftplug};
      \draw (4.3,6.25) node {{\scriptsize remote client}};
      \draw[-, thick] (3,6.75) -- (4,6.75);
      \draw (5.5,6.75) node {{\scriptsize port 80}};
      
      \draw (1.5,4.5) node {{\scriptsize AhYvJj737BtY24\ldots}};

      \draw [fill=lightred,thick] (4,4.85) rectangle +(4,0.8);
      \draw (5.5,5.25) node {{\scriptsize GaWKnFs1s2syow\ldots}};
      \draw (6,5.9) node {{\scriptsize local miner}};
      \draw (7.4,5.25) node {\ssd};
      \draw[-, thick] (3,5.25) -- (4,5.25);

      \draw (4.3,4.1) node {\leftplug};
      \draw (4.3,3.6) node {{\scriptsize remote miner}};
      \draw[-, thick] (3,4.1) -- (4,4.1);
      \draw (5.6,4.1) node {{\scriptsize port 8026}};

      \draw [fill=lightred,thick] (4,2) rectangle +(4,0.8);
      \draw (6,3.1) node {{\scriptsize local miner}};
      \draw (5.5,2.4) node {{\scriptsize FbXZjuFsZvQ1Ae\ldots}};
      \draw (7.4,2.4) node {\ssd};
      \draw[-, thick] (3,2.4) -- (4,2.4);

      \draw[<->, thick, dashed] (1.5,2.5) -- (1.5,4);
      \draw (0.8,2.7) node {{\scriptsize port 8030}};
      \draw (0.8,3.7) node {{\scriptsize port 8032}};

      \draw (4.3,1.1) node {\leftplug};
      \draw (4.3,0.6) node {{\scriptsize remote miner}};
      \draw[-, thick] (3,1.1) -- (4,1.1);
      
      \draw (4.3,-0.2) node {\leftplug};
      \draw (4.3,-0.7) node {{\scriptsize remote client}};
      \draw[-, thick] (3,-0.2) -- (4,-0.2);
      \draw (5.6,-0.2) node {{\scriptsize port 8001}};

      \draw [fill=verylightgray,dashed] (6.3,0.2) rectangle +(6,1.6);
      \draw[<->, thick, dashed] (4.8,1.1) -- (6.4,1.1);
      \draw (5.6,1.4) node {{\scriptsize port 8025}};
      \draw[-, thick] (7.15,1.1) -- (8.1,1.1);
      \draw [fill=lightred,thick] (8.2,0.7) rectangle +(4,0.8);
      \draw (6.8,1.1) node {\rightplug};
      \draw (10.1,0.4) node {{\scriptsize local miner in our machine}};
      \draw (9.5,1.1) node {{\scriptsize our public key}};
      \draw (11.6,1.1) node {\ssd};
    \end{tikzpicture}
  \end{center}
  \caption{The network architecture of a couple of Hotmoka nodes based on Mokamint.}\label{fig:hotmoka_mokamint_network}
\end{figure}

Fig.~\ref{fig:hotmoka_mokamint_network} shows the network architecture of a couple
of Hotmoka nodes based on Mokamint, that are two peers of the same blockchain. The two Mokamint instantiations
communicate between each other, in order to exchange blocks and synchronize their copy of the blockchain.
Each Mokamint instantiation runs an instance of Hotmoka, that specifies the semantics of the transactions
that can be executed on the blockchain. Each peer exports the node interface (Fig.~\ref{fig:node_hierarchy}),
respectively, in this example, at port 80 and 8001. Remote clients, such as \texttt{moka} or Mokito, can connect to such
ports and run transactions on the nodes, that will be automatically executed in all peers.
Fig.~\ref{fig:hotmoka_mokamint_network} shows that each peer runs a local miner\index{miner!local}, that is, a software
tool that finds nonces on disk in order to create the next blocks. It is well possible to have peers
without any miner, in which case they just synchronize their copy of the blockchain but do not contribute
new blocks. However, for simplicity, our docker images will always provide a local miner to a node.
As the same figure shows,
some peer might allow the connection of \emph{remote miners}\index{miner!remote},
that is, miners, identical to the local ones,
but running as a remote network service. This allows peers to exploit the willingness of remote users
to provide mining power to them by just plugging remote miners, without running a full node. If such remote miners
contribute to the creation of new blocks,
the prize for this creation will be shared between the peer and the remote miners.

\subsection{Interacting with a Hotmoka node based on Mokamint}\label{subsec:interacting_mokamint}

The software components of Fig.~\ref{fig:architecture} are normally exposed through some IP ports, so that
remote users can interact with them. Interaction with Hotmoka is possible through the \texttt{moka}\index{moka@{\texttt{moka}}} command-line utility.
Interaction with Mokamint's peer and miner is possible through a few utilities of Mokamint itself:
%
\begin{itemize}
\item \texttt{moka} allows interaction with any Hotmoka node, independently from the underlying engine (Mokamint or Tendermint);
\item \texttt{mokamint-node}\index{mokamint-node@{\texttt{mokamint-node}}} allows one to interact with the Mokamint node;
\item \texttt{mokamint-miner}\index{mokamint-miner@{\texttt{mokamint-miner}}} allows one to interact with the miners of a Mokamint node;
\item \texttt{mokamint-application}\index{mokamint-application@{\texttt{mokamint-application}}} allows one to interact with the application of a Mokamint node, including Hotmoka.
\end{itemize}
%
While \texttt{mokamint-application} implements very generic operations, valid for all applications running on top of
Mokamint, \texttt{moka} implements operations specific to Hotmoka only.

Let us start by using some of these utilities with the Mokamint-based main network \texttt{\serverMokamintMainChainId{}}
of Hotmoka. Fig.~\ref{fig:endpoints} reports the endpoints of a node of that network: Hotmoka node, Mokamint node and a remote miner
of that node.

First of all, you need to enter the docker container, where all needed utilities are already installed:
%
\inputCommand{docker_mokamintnode_enter_container}
%
You should see the prompt of docker now, something like \texttt{hotmoka@d8bf8010d847:~\$}.
The following commands must be executed inside that running container.

The first command lets us see which are the peers known by a node:
%
\index{mokamint-node@{\texttt{mokamint-node}}!{peers ls}@{\texttt{peers ls}}}
\inputCommand{mokamintnode_peers_ls}
%
Note that the command above is talking to Mokamint, not to Hotmoka: it is Mokamint
that takes care of networking. Moreover, the command is run inside a docker container, so
that you do not need to install anything in your machine. Later we will see how to
run it, directly, in your machine. In any case, the output will be something like:
%
\inputOutput{mokamintnode_peers_ls}
%
These are the peers connected to node \serverMokamintMainPublic{} of the main network.
You can try to replace the URI in the command with that of one of such peers,
to see its peers.

Let us check which miners are connected to \serverMokamintMainPublic{}:
%
\index{mokamint-node@{\texttt{mokamint-node}}!{miners ls}@{\texttt{miners ls}}}
\inputCommand{mokamintnode_miners_ls}
%
The output tells us that there are two miners: a local and a remote one:
%
\inputOutput{mokamintnode_miners_ls}
%
The local miner is working in the same machine of the Mokamint node, connected directly to it.
The remote miner, instead, is just a plug, listening to port $8025$: any external miner
can connect to that port and help peer \serverMokamintMainPublic{} to create new blocks.
Whenever this help actually leads to the creation of a new block, the reward for that block
will be split between the helper and the node at \serverMokamintMainPublic{}. We will show
soon how to start our own miner that helps an existing Mokamint node.

Let us query the configuration of the Mokamint node now:
%
\index{mokamint-node@{\texttt{mokamint-node}}!{config show}@{\texttt{config show}}}
\inputCommand{mokamintnode_config_show}
%
The output reports all consensus parameters of Mokamint. They are the same for all nodes
belonging to the same network \texttt{\serverMokamintMainChainId{}}:
%
\inputOutput{mokamintnode_config_show}
%
Among the data reported above, note the expected chain id of the main network
(\texttt{\serverMokamintMainChainId{}}) and the average time between two successive blocks
($20000$ milliseconds, \ie{}, $20$ seconds). This is an average because Mokamint
has probabilistic finality.

We can list the last five blocks of the blockchain:
%
\index{mokamint-node@{\texttt{mokamint-node}}!{chain ls}@{\texttt{chain ls}}}
\inputCommand{mokamintnode_chain_ls}
%
The output, for each of the last five blocks, reports block's hash and time of creation.
Moreover, each block is associated with two public keys, for the Mokamint node and for the miner
that created the block and shared the reward for its creation:
%
\inputOutput{mokamintnode_chain_ls}
%
It is also possible to have a detailed description of a block such as, for instance, the head of the
blockchain:
%
\index{mokamint-node@{\texttt{mokamint-node}}!{chain show}@{\texttt{chain show}}}
\inputCommand{mokamintnode_chain_show_head}
%
Among the technical information contained in the block, it is possible to recognize
the \emph{value} of the deadline, which is inversely proportional to the mining power
of the blockchain as a whole, the state of the \emph{application} (in this network: of Hotmoka)
reached at the end of this block and the user requests included and executed in the block, if any:
%
\inputOutput{mokamintnode_chain_show_head}

\subsection{Start a remote miner for an existing node}\label{subsec:mining_hotmoka_mokamint}

The simplest scenario is that of a user who wants to provide mining power to an existing node of a blockchain
based on Mokamint, such as \serverMokamint{}. This user does not want to run a full node of the blockchain,
for instance because he does not want to perfom a long synchronization, nor hold a copy of the full blockchain
in his machine, or maybe because his internet connection is not stable enough for running a full node, or
just because he does not want to keep his machine always on, day and night. This user can contribute
to the creation of new blocks, and earn crypto for this, by running a remote miner
for an existing Hotmoka node based on Mokamint (Fig.~\ref{fig:hotmoka_mokamint_network}),
which can be done in two ways. The simplest is to run the
Mokaminter Android app~\cite{mokaminter_google_play} on a mobile phone or tablet. This should be so
simple that we do not really describe it here: just install the app and try to start mining. The other possibility
is to mine on a local machine, and this is what we describe it here. In both cases, the miner is identified by a key pair,
that is where the crypto earned for mining gets sent. The key is created automatically by Mokaminter,
while it must be created manually when mining on a local machine.

There is a docker image for running a remote miner for Mokamint~\cite{mokamint_dockerhub}.
You first need to create a key pair that will identify you as a miner of the blockchain, then create a mining configuration,
compatible with the Mokamint node that you are going to mine for, that includes a plot file for mining.
The larger this plot file, the more likely it is that the miner will be selected for creating the next blocks
of the blockchain and more coins will be earned by your miner, on average.
For instance, in order to mine for the Mokamint node running the Hotmoka testnet
and published at \serverMokamintMining{}, you start with the creation of the key pair
that will identify you as a miner:
%
\inputCommand{docker_create_keys_for_mining}
\inputOutput{docker_create_keys_for_mining}
%
\begin{commentbox}
We create the key pair inside a docker container above and then export it outside.
Thus is because, in this chapter, we perform everything with the docker tool only.
You can of course do the same by invoking \texttt{moka keys create} directly if you
have installed on your local machine (Sec.~\ref{subsubsec:moka_local}).
\end{commentbox}
%
The command above asks for a password, that you will use to control the key pair,
it creates the key pair in your local file system, named \texttt{miner.pem},
and prints its public key.
Write down the password on paper, since it is not possible to recover it if lost.
Save \texttt{miner.pem} in a safe place as well.

You can create the mining configuration now:
%
\index{docker!docker config\_miner@{\texttt{config\_miner}}}
\inputCommand{docker_config_miner}
%
You can replace 5000 with any other positive number of nonces in order to increase or decrease the size of the plot file.
Similarly, you can replace \texttt{\serverMokamintMining{}} with
\texttt{\serverMokamintMainMining{}} if you want to mine for the mainnet of Hotmoka instead of the testnet.

After a mining configuration has been created, you can start mining:
%
\index{docker!docker mine@{\texttt{mine}}}
\inputCommand{docker_mine}
%
You should see on the screen the flow of the logs of the miner.
You can terminate the miner by pressing the ENTER key.
Or you can leave the miner working in the background
by pressing CTRL+p followed by CTRL-q, as always with docker.

You can stop an already running miner by simply stopping its docker container (\texttt{docker stop miner}).
You can restart a stopped miner by running the above \texttt{mine} command again.
You normally do not recreate the mining configuration when you restart a miner: its creation is a one-shot action.
Nevertheless, you can recreate the configuration if, for instance, you want to change the size of the plot file.
For that, stop your miner (if running) and just rerun the docker command for \texttt{config-miner}.
Once the mining configuration has been recreated, you can restart your miner with \texttt{mine}.

\begin{commentbox}
If you reconfigure the miner by rerunning \texttt{config-miner} (for instance, to make its plot larger), it is
convenient to reuse the same \texttt{miner.pem}, to avoid having many keys around and
consolidate all your earnings on the same key.
\end{commentbox}

Your miner will earn coins for the mining activity. You can check the current balance of the public key
that identifies your miner, as follows:
%
\index{docker!docker balance@{\texttt{balance}}}
\inputCommand{docker_check_balance}
%
When the balance will finally be non-zero, you can bind the public key to the actual account that holds its crypto coins, with
%
\index{moka@{\texttt{moka}}!keys bind@{\texttt{keys bind}}}
\inputCommand{moka_keys_bind_miner}
%
that requires you to have \texttt{moka} installed in your machine (Sec.~\ref{subsubsec:moka_local})
and to be run in the directory where you have saved the key pair file \texttt{miner.pem}.
Note that the above command will fail if your miner has not yet created any block. Therefore,
you must probably wait for some time before running \texttt{moka keys bind}. When successful,
the \texttt{miner.pem} will be renamed into the address of the account inside the state of
Hotmoka, something like \texttt{\gameteShort{}.pem}.

\subsection{Key pairs and accounts}\label{subsec:keypairs_and_accounts}

The way Hotmoka deals with keys and accounts is different from that of other blockchains:
%
\begin{itemize}
\item a key pair is the private key/public key pair that can be used to identify a miner, or
	a node, or to control an account. It is uniquely identified by a \texttt{.pem} file and
	its associated password (if any). Equivalently, it is uniquely identified by $12$ words and
	(maybe) a password. This means that you can \emph{move} a key pair by simply copying its
	\texttt{.pem} around across machines, or you can export the \texttt{.pem} file into its
	$12$ words (\texttt{moka keys export file.pem}\index{moka@{\texttt{moka}}!keys export@{\texttt{keys export}}})
	and import it elsewhere from those $12$ words
	(\texttt{moka keys import <12 space-separated words here> --name file.pem}\index{moka@{\texttt{moka}}!keys import@{\texttt{keys import}}});
\item an account is an actual Java object in the state
	of Hotmoka: it has a public key and can be controlled with a key pair matching that public key.
	It follows that more Java account objects can be controlled with the same key pair,
	if they all use the same public key: they will have distinct identities and will keep
	different crypto amounts, but they can be controlled with the same key pair.
	An account is needed only inside Hotmoka, it is never
	used by Mokamint or Tendermint. Namely, you need an account if you want to send crypto across
	accounts or, in general, if you want to run a Hotmoka transaction (install a smart contract
	and run its methods).
\end{itemize}
%
The natural question, now, is how we can identify an account: we need its key pair
(a \texttt{.pem} file) \emph{and} the address where the account object is allocated in the state
of Hotmoka (something like \texttt{\gameteShort{}}). Hotmoka's solution is
to identify an account with a \texttt{.pem} file whose name is exactly the address of the
account, for instance: \texttt{\gameteShort{}.pem}.

It turns out that key pairs, if used to identify a miner, a node or the gamete of the network
(Sec.~\ref{subsec:start_hotmoka_mokamint}), automatically have a corresponding account
that can be used to manipulate the earned crypto. In order to find out which is that account,
we can use the \texttt{moka keys bind}\index{moka@{\texttt{moka}}!keys bind@{\texttt{keys bind}}} comand, as seen above:
%
\inputCommand{moka_keys_bind_miner}
%
(replace the URI above with \texttt{\serverMokamintMain{}} if you are mining for the mainnet of Hotmoka).
This will rename \texttt{miner.pem} into something like \texttt{\gameteShort{}.pem}, that you can
use to take control of the crypto earned for mining, send it to other users or spend it
to install and run smart contracts. If you want to be able to control the account on another machine,
you can just copy this \texttt{\gameteShort{}.pem} file to the other machine,
or you can export the account into $36$ words (\texttt{moka accounts export \gameteShort{}})
and import it in the other machine (\texttt{moka accounts import <36 space-separated words here>}).

Information about balance and public key of an account is public and can be accessed
without holding its controlling \texttt{.pem} file\index{moka@{\texttt{moka}}!accounts show@{\texttt{accounts show}}}:
%
\inputCommand{moka_accounts_show}
\inputOutput{moka_accounts_show}
%
Of course, you need the \texttt{\gameteShort{}.pem} file in your local machine if you want to use
the crypto of that account (and you must know its password, if any). For instance, if you want to move $1000$
coins from account \texttt{\gameteShort{}} to account \texttt{\manifestShort{}}, you can write
%
\inputCommand{moka_accounts_send_from_gamete}\index{moka@{\texttt{moka}}!accounts send@{\texttt{accounts send}}}
%
That requires that you can control the sender (you need \texttt{\gameteShort{}.pem}),
but not that you control the receiver (you do not need \texttt{\manifestShort{}.pem}).

\subsection{Start a node of an existing blockchain}\label{subsec:join_hotmoka_mokamint}

This is the most typical situation. Namely, this occurs if you want to join an existing blockchain,
with a node that mines new blocks and receives blocks created by the other peers.
This docker image provides a script for this situation. This script includes the creation of
a local miner as well, or otherwise your node would not be able to mine new blocks.
Because of this, the scripts deal with two key pairs: the former identifies the node,
and is kept inside the machine running the script, and the latter identifies the miner,
can be stored elsewhere and only its public key is needed here.

The process is consequently split in two:
%
\begin{itemize}	
\item configure the node (\texttt{config-clone});
\item run the node (\texttt{go}).
\end{itemize}
%
Each phase is the execution of a script inside this docker image. The script
\texttt{config-clone} is meant to be run only once, while \texttt{go} can be run, stopped and run again,
whenever you want to start, stop and restart a node. You can also pause it and unpause it.
The reason for splitting the process in two scripts is that it allows one
to manually edit the configuration created by \texttt{config-clone} before running the node,
although we will not show this here. Moreover, having distinct scripts allows \texttt{go}
to be stopped and run again, repeatedly, whenever you want to stop and restart a node.

The following instructions assume that you have a reliable internet connection.
If the connection is too slow, or flickering, or if it disconnects for some time,
the synchronization of the node will likely fail. In that case, stop and restart \texttt{go} later, hoping
to complete the synchronization, eventually. Please consider that the synchronization of the node
requires to download all the history of the blockchain that you are joining, which might take time
(hours, days, even weeks, depending on the age of the blockchain and the quality of the internet
connection).

\subsubsection{Configure the node (\texttt{config-clone})}\label{subsubsec:mokamint_config_clone}

The \texttt{config-clone} script creates the configuration directory of a node
that joins an existing Hotmoka blockchain based on the Mokamint proof of space engine.
For that, you must specify the URI of a node of this blockchain, from where
the configuration information will be fetched. This is important, since all nodes of the
same blockchain must have the same consensus parameters.

The first thing to do is to create a key pair for the miner of the new node
that you want to start. You can do this as in Sec.~\ref{subsec:mining_hotmoka_mokamint},
hence generating \texttt{miner.pem}.

You can then run the script that configures the node. Use for it the base58-encoded public key
of the key pair that you have created above. Specify \serverMokamintPublic{}
as the URI of a node of the blockchain to join.
Specify the size of the plot file to use for the proof of space:
the larger, the more blocks will be created by your node, but also more disk space
will be allocated for mining. We will use two volumes: \texttt{chain} will contain
the actual blockchain data and \texttt{hotmoka\_mokamint} will contain the configuration
information created for the node. By using volumes, we can share that information
across successive invocations of docker:
%
\inputCommand{docker_mokamint_config_clone}
%
Note that the script above will create another key pair (and ask you about the relative password),
that will be kept \emph{inside} the container. This key pair identifies the node
and will be used to sign the blocks that the node will create. Because of this, that key
must remain inside the container.

\subsubsection{Run the node (\texttt{go})}\label{subsubsec:mokamint_go}

After configuring the node, you can run it with the go script:
%
\inputCommand{docker_mokamint_go}
%
The command above will start a node of the same blockchain of \serverMokamint{},
that allows connections to the ports:
%
\begin{description}
\item[8001:] this is the port where the Hotmoka node is published by default;
	it can be used to contact the node, install and run smart contracts;
\item[8025:] this is the port where mining services can connect by default;
	mining services help your node produce new blocks; note that this requires to open
	a remote miner in your node, listening at port 8025 (\texttt{mokamint-node miners add});
\item[8030:] this is the port where Mokamint can be reached for public queries, by default;
\item[8031:] this is the port where Mokamint can be reached for restricted operations, by default;
	note the we have restricted its access to localhost only,
	since we do not want our Mokamint node to be freely reconfigured remotely.
\end{description}
%
After the command above, you should see that the node will start up and begin synchronizing
from \serverMokamintPublic{}. This will take some time (hours, days, weeks\ldots) depending
on the age of the cloned blockchain and on the speed of your internet connection.
You can leave the container in the backrground by entering CTRL+p, CTRL+q, as always in docker.

You can then monitor the progress of the synchronization by entering the running container
and executing the \texttt{mokamint-node} command:
%
\begin{shellcommandbox}\begin{ttlst}
docker exec -it hotmoka /bin/bash
\end{ttlst}\end{shellcommandbox}
%
and then inside the container:
%
\begin{shellcommandbox}\begin{ttlst}
hotmoka@e41eda9afd3b: mokamint-node chain ls 10
\end{ttlst}\end{shellcommandbox}
%
You can also see the manifest of the node, that is identical to that of any other node of the joined blockchain:
%
\begin{shellcommandbox}\begin{ttlst}
hotmoka@e41eda9afd3b: moka nodes manifest show
\end{ttlst}\end{shellcommandbox}
%
Remember that you can always check the logs of a running docker container, for instance with:
%
\begin{shellcommandbox}\begin{ttlst}
docker logs -f hotmoka
\end{ttlst}\end{shellcommandbox}
%
You can also copy the key pair of the node outside the container, to keep it in a safe place:
%
\begin{shellcommandbox}\begin{ttlst}
docker cp hotmoka:/home/hotmoka/hotmoka_mokamint/node.pem .
\end{ttlst}\end{shellcommandbox}
%
We suggest that you always keep a copy of \texttt{node.pem} and \texttt{miner.pem} in a safe
place and write down their passwords, if any.

Finally, you can pause, unpause and stop the node with the usual Docker commands
\texttt{docker pause hotmoka}, \texttt{docker unpause hotmoka} and
\texttt{docker stop hotmoka}. After a stop, you can execute the \texttt{go} script
again if you want to resume the node.

\begin{commentbox}
If you want to reconfigure the node, for instance because you want to resize its plot file,
you can stop the node (\texttt{docker stop hotmoka}), rerun the \texttt{config-clone} script
and then the \texttt{go} script. Note however that, by default, \texttt{config-clone} will
create a \emph{new} \texttt{node.pem} key to identify the node and will
erase the directory where the blockchain is kept. Therefore, you will end up with two
node keys (the old one and the new one), while it would be normally easier to recycle the old key
and consolidate all earnings in that same key. Moreover, the chain directory will have to be
downloaded again, from scratch, with a long synchronization, when \texttt{go} will be invoked.
Because of this, we suggest that, when rerunning the \texttt{config-clone} script, you set
the environment variables \texttt{-e KEEP\_NODE\_PEM=true} and \texttt{-e KEEP\_CHAIN=true}
in the docker invocation.

\end{commentbox}

\subsection{Start the first node of a new blockchain}\label{subsec:start_hotmoka_mokamint}

This situation is much rarer. It occurs when you want to start a brand new blockchain from scratch,
by minting its genesis block and initializing its store. Later, other nodes can join the new blockchain
with the technique described in Sec.~\ref{subsec:join_hotmoka_mokamint}.

The docker image of Hotmoka provides scripts for starting a brand new blockchain.
These scripts include the creation of a local miner as well, or otherwise your node
would not be able to mine new blocks. Because of this, the scripts deal with two key pairs:
the first identifies the node, and is kept inside the machine running the script,
and the second identifies the miner, can be stored elsewhere and only
its public key is needed for starting the node.

The process is consequently split in three:
%
\begin{itemize}
\item configure the node (\texttt{config-new})
\item initialize the node (\texttt{init})
\item run the node (\texttt{go})
\end{itemize}
%
Each phase is the execution of a script inside the docker image. The scripts \texttt{config-new}
and \texttt{init} are meant to be run only once, while \texttt{go} can be run, stopped and run again,
whenever you want to start or stop a node. You can also pause it and unpause it.
The reason for splitting the process in three scripts is that it allows one to manually edit
the configuration created by \texttt{config-new} before initializing and running the node,
although we will not show this here. Moreover, having distinct scripts allows \texttt{go}
to be stopped and run again, repeatedly, whenever you want to stop and restart a node.

\subsubsection{Configure the node (\texttt{config-new})}\label{subsubsec:mokamint_config_new}

The configuration requires the creation of a brand new key pair that will identify the miner of the node.
You can do this as in Sec.~\ref{subsec:mining_hotmoka_mokamint},
hence generating \texttt{miner.pem}.
Moreover, you will need another key pair, for the gamete\index{gamete} account.
This is an account of Hotmoka that holds all cryptocurrency minted at start-up (if any).
It can also be used as faucet of the node, if your node allows a free faucet.
In any case, you need a key pair for the gamete account as well.
Therefore, follow the instructions in Sec.~\ref{subsec:mining_hotmoka_mokamint} and create
\texttt{gamete.pem} as well:
%
\inputCommand{docker_create_keys_for_mokamint_gamete}
\inputOutput{docker_create_keys_for_mokamint_gamete}
%
You can now configure the node, by specifying the public key of the miner and that of the gamete:
%
\inputCommand{docker_mokamint_config_new}
%
Note that the public key of the miner is reported in base58,
while that of the gamete is reported in base64, currently.
The target block creation time, in milliseconds, is the average time between the creation of two successive blocks.
The chain identifier identifies the new network and must be used, for instance,
in the transaction requests sent to the Hotmoka nodes of the network.
The script above will prompt for the password of the key pair used for signing the new blocks, that it creates.
Enter your chosen password or leave it blank. The script will configure the node and create a plot file for its miner.

\subsubsection{Initialize the node (\texttt{init})}\label{subsubsec:mokamint_init_node}

The initialization of the node consists in the execution of a few initial transactions
that create the genesis block, the manifest and the gas station of the node. You can do this with:
%
\inputCommand{docker_mokamint_init}
%
This will take a few seconds. You should see the logs of the executed transactions,
until the script terminates.

\subsubsection{Run the node (\texttt{go})}\label{subsubsec:mokamint_go_new_node}

Once the node has been configured and initialized, you can run it.
Follow for this the instructions reported in Sec.~\ref{subsec:join_hotmoka_mokamint}
for the \texttt{go} script.

\subsection{Migrate a node}\label{subsec:migrate_a_node}

What identifies a node is its key pair, used to sign the blocks that it creates.
If you ran either \texttt{config-clone} or \texttt{config-new}, these scripts will
have created a \texttt{node.pem} key pair file inside the configuration directory of the docker
container, to identify the node. It might happen that you want to
\emph{migrate} that node, \ie{}, you might want to move it into another machine, for instance
because it has a larger disk space or a better internet connection. In order to keep the
identity of the node, you need to recycle the same \texttt{node.pem} that was used in
the previous machine. But running \texttt{config-clone} in the new machine will created
a \emph{new} \texttt{node.pem} and you will end up with two key pairs, that of the old
node and that of the new node. This is typically problematic, since you will have some coins collected
while mining with the old key pair and other coins collected while mining with the new key pair.
Instead, you normally want to keep the old \texttt{node.pem} and consolidate all earnings
into that same key pair.

Assume then that you have copied the key pair \texttt{node.pem} of the old node in your working directory of the new machine.
Remember that you can extract it from the old node, running in the old machine, with
%
\begin{shellcommandbox}\begin{ttlst}
docker cp hotmoka:/home/hotmoka/hotmoka_mokamint/node.pem .
\end{ttlst}\end{shellcommandbox}
%
In the new machine, create the container, but do not start it yet:
%
\inputCommand{docker_mokamint_create_config_clone}
%
Note that we have given a name to the container (\texttt{hotmoka}) and that we asked to keep the existing
\texttt{node.pem} key pair. Moreover, we are always using the same public key for the local miner of the new node as well,
so that we also consolidate all earnings for the local miner. After creating the container in the new machine, let us
inject the desired \texttt{node.pem} inside it:
%
\begin{shellcommandbox}\begin{ttlst}
docker cp node.pem hotmoka:/home/hotmoka/hotmoka_mokamint/
\end{ttlst}\end{shellcommandbox}
%
We can start the container \texttt{hotmoka} now: it will use the injected \texttt{node.pem}, instead
of creating a new one, since we used \texttt{-e KEEP\_NODE\_PEM=true} when we created that container above:
%
\begin{shellcommandbox}\begin{ttlst}
docker start --attach -i hotmoka
\end{ttlst}\end{shellcommandbox}
%
This will start the container and create the configuration. When it will complete, you can run the migrated node, in the new machine,
with the \texttt{go} script. It will perform an initial synchronization and then start mining,
with the same identity as the old node.

\section{Hotmoka blockchains based on Tendermint (proof of stake)}\label{sec:hotmoka_tendermint}

Tendermint~\cite{Tendermint}\index{Tendermint}, now Ignite, is a
Byzantine-fault tolerant engine for building blockchains, that
replicates a finite-state machine on a network of nodes across the world.
The finite-state machine is often referred to as a \emph{Tendermint app}.
The nice feature of Tendermint is that it takes care of all
issues related to networking and consensus, leaving to the
developer only the task to develop the Tendermint app.

\begin{figure}[th]
  \begin{center}
    \begin{tikzpicture}[scale=0.7]
      \draw [fill=lightviolet,thick] (0,0) rectangle +(10,11);
	  \draw (5,10.2) node {Hotmoka node};
	  
	  \draw [fill=lightviolet,thick] (15,0.5) rectangle +(8,5);
	  \draw (19,4.9) node {other Hotmoka node};
	  \draw[<->,thick,dashed] (9,2) -- (17,2);
	  \draw (12.1,2.6) node {port 26656};
	  \draw[<-,thick,dashed] (17,3.8) .. controls (11,5.5) .. node[near start,sloped,below] {\small port 8002} (17,9);
	  \draw[<->,thick,dashed] (3.5,3) -- (3.5,5.5);
	  \draw (6.5,4.2) node {ports 26657, 26658};

	  \draw [fill=yellow,thick] (17,1.5) rectangle +(4,1);
	  \draw (19,2) node {Tendermint};

	  \draw[<->,thick,dashed] (19,2.5) -- (19,3.2);

	  \draw [fill=yellow,thick] (17,3.2) rectangle +(4,1);
  	  \draw (19,3.7) node {Hotmoka};

  	  \draw [fill=lightblue,thick] (15,7) rectangle +(8,2);
	  \draw (19,8) node {clients (moka, Mokito,\ldots)};

	  \draw [fill=yellow,thick] (1,5.5) rectangle +(8,3.5);
	  \draw (5,8.5) node {Hotmoka};
	  \draw (5,6.8) node {
		$\left(\begin{array}{c}
		\text{handles transactions}\\
		\text{keeps state on Xodus}\\
		\text{interprets Takamaka}
		\end{array}\right)$
	  };
	  \draw[<-,thick,dashed] (9,8.4) -- (15,8.4);
	  \draw (12,9) node {port 8002};

	  \draw [fill=yellow,thick] (1,1) rectangle +(8,2);
	  \draw (5,2.5) node {Tendermint};
	  \draw (5,1.6) node {(consensus, networking)};
    \end{tikzpicture}
  \end{center}
  \caption{The network architecture of the Hotmoka node based on Tendermint.}\label{fig:hotmoka_tendermint_network}
\end{figure}

There is a docker image that implements a Hotmoka node as a Tendermint app
for programming in Takamaka over Tendermint. We have already used that node
previously, since that installed at
\serverTendermint{} is a node of that type.
Fig.~\ref{fig:hotmoka_tendermint_network}
shows the architecture of a Tendermint Hotmoka node.
It consists of a few components.
The Hotmoka component is the Tendermint app that
implements the transactions on the state, that is the installation
of jars and the execution of code written in the Takamaka subset of Java. This part is the same in every
implementation of a Hotmoka node, not only for this one based on Tendermint.
The database that contains the state is implemented by
using the Xodus transactional database by IntelliJ.
What is specific here, however, is that transactions are put inside a blockchain
implemented by Tendermint. The communication occurs, internally, through the two TCP ports
26657 and 26658, that are the standard choice of Tendermint for communicating with an app.
Clients can contact the Hotmoka node
through any port, typically but not exclusively~8001 or~8002, as a websocket service.
The node can live alone but is normally integrated with other Hotmoka nodes based on Tendermint, so that
they execute and verify the same transactions, reaching the same state at the end. This happens through
the TCP port 26656, that allows Tendermint instances to \emph{gossip}:
they exchange transactions and information on peers and finally reach consensus.
Each node can be configured to use a different port to communicate with clients,
which is useful if, for instance, ports 8001 or 8002 (or both)
are already used by some other service.
Port 26656 must be the same for all nodes in the network, since they must communicate on
a standard port.

As shown in Fig.~\ref{fig:hotmoka_tendermint_network}, clients connect to the Hotmoka
part of any node of the blockchain. The exact node is irrelevant, since all nodes
of the same blockchain are kept synchronized by the underlying Tendermint engine.
For instance, we can use \serverTendermint{} to play with
accounts and Takamaka contracts. However, we might want to
install our own node instead, part of the same
blockchain network of \serverTendermint{}
or part of a brand new blockchain. In the former case, our own node will execute
the same transactions of \serverTendermint{}, so that we can be sure that they are
executed according to the rules. In the latter case, we can have our own blockchain that
executes our transactions only, instead of using a shared blockchain such as that
at \serverTendermint{}.

\subsection{Start a node of an existing blockchain}\label{subsec:join_hotmoka_tendermint}

As for Hotmoka nodes based on Mokamint~\ref{subsec:join_hotmoka_mokamint}, this task is split in two scripts,
\texttt{config-clone} and \text{go}.

\subsubsection{Configure the node (\texttt{config-clone})}\label{subsubsec:tendermint_config_clone}

The \texttt{config-clone} script creates the configuration directory of a node
that joins an existing Hotmoka blockchain based on the Tendermint byzantine
consensus engine. For that, you must specify the URI of a node of this blockchain,
from where the configuration information will be fetched, such as
\serverTendermint{}. This is important, since all nodes of the
same blockchain must have the same consensus parameters.
You must also specify the average block creation rate of the blockchain,
that in the case of \serverTendermint{} is $10,000$ milliseconds.
We will use two volumes: \texttt{chain} will contain the actual blockchain data
and \texttt{hotmoka\_tendermint} will contain the configuration information
created for the node. By using volumes, we can share that information
across successive invocations of docker.

You can then run the script that configures the node.
%
\inputCommand{docker_tendermint_config_clone}
%
The script above will create a key pair with empty password, that will be kept inside the container.
This key pair identifies the node and will be used to sign the blocks that the node will create,
if it will ever become a validator of the network. It must remain inside the container,
although you may want to extract a copy from the container to your local host.

\subsubsection{Run the node (\texttt{go})}\label{subsubsec:tendermint_go}

After configuring the node, you can run it with the go script:
%
\inputCommand{docker_tendermint_go}
%
The command above will start a node of the same blockchain of \serverTendermint{},
that allows connections to the ports:
%
\begin{description}
\item[8001:] this is the port where the Hotmoka node is published by default;
	it can be used to contact the node, install and run smart contracts;
\item[26656:] this is the port where the Tendermint engine communicates with its peers.
\end{description}
%
After the command above, you should see that the node will start up and begin synchronizing
from \serverTendermint{}. This will take some time (hours, days, weeks\ldots) depending
on the age of the cloned blockchain and on the speed of your internet connection.
You can leave the container in the backrground by entering CTRL+p, CTRL+q, as always in docker.

You can then enter the running container and check the manifest of the node,
that is identical to that of any other node of the joined blockchain:
%
\begin{shellcommandbox}\begin{ttlst}
docker exec -it hotmoka /bin/bash
\end{ttlst}\end{shellcommandbox}
%
and then inside the container:
%
\begin{shellcommandbox}\begin{ttlst}
hotmoka@e41eda9afd3b: moka nodes manifest show
\end{ttlst}\end{shellcommandbox}
%
Remember that you can always check the logs of a running docker container, for instance with:
%
\begin{shellcommandbox}\begin{ttlst}
docker logs -f hotmoka
\end{ttlst}\end{shellcommandbox}

The resulting node will not be a validator. For that, you need to buy some validation power
from a validator who is willing to sell it to you.
Check the \texttt{moka nodes tendermint validators}\index{moka@{\texttt{moka}}!nodes tendermint validators@{\texttt{nodes tendermint validators}}} commands.

\subsection{Start a node of a new blockchain}\label{subsec:start_hotmoka_tendermint}

As in Sec.~\ref{subsec:start_hotmoka_mokamint}, this task is split in three scripts,
\texttt{config-new}, \texttt{init} and \texttt{go}.

\subsubsection{Configure the node (\texttt{config-new})}\label{subsubsec:tendermint_config_new}

You will need a key pair, for the gamete\index{gamete} account.
This is an account of Hotmoka that holds all cryptocurrency minted at start-up (if any).
It can also be used as faucet of the node, if your node allows a free faucet.
You can follow the instructions in Sec.~\ref{subsec:mining_hotmoka_mokamint} and create
\texttt{gamete.pem} as well:
%
\inputCommand{docker_create_keys_for_tendermint_gamete}
\inputOutput{docker_create_keys_for_tendermint_gamete}
%
You can now configure the node, by specifying the public key of the gamete:
%
\inputCommand{docker_tendermint_config_new}
%
Note that the public key of the gamete is reported in base64, currently.
The target block creation time, in milliseconds, is the average time between
the creation of two successive blocks. The chain identifier identifies
the new network and must be used, for instance, in the transaction
requests sent to the Hotmoka nodes of the network.

\subsubsection{Initialize the node (\texttt{init})}\label{subsubsec:tendermint_init_node}

The initialization of the node consists in the execution of a few initial transactions
that create the genesis block, the manifest and the gas station of the node. You can do this with:
%
\inputCommand{docker_tendermint_init}
%
This will take a few seconds. You should see the logs of the executed transactions,
until the script terminates.

\subsubsection{Run the node (\texttt{go})}\label{subsubsec:tendermint_go_new_node}

Once the node has been configured and initialized, you can run it.
Follow for this the instructions reported in Sec.~\ref{subsec:join_hotmoka_tendermint}
for the \texttt{go} script.

\subsection{Manifest and validators}\label{subsec:manifest_and_validators}

The information reported by \texttt{moka nodes manifest show} refers to two accounts that have been
created during the initialization of the node:
%
\begin{shellbox}\begin{ttlst}
gamete: fcc6ad9a4cd4109dcb554df6f1d951f770d42cdb4c2caeb8fa713c1d4189b4fa#0
  balance: 1000000000000000000000000000000000
  maxFaucet: 0

validator #0: b437832a688dbb89b145d380b7cb3eb841d7cb09fb600d27be43cd670e8b43f9#0
  id: 030203B36BA4EDF0182D7D40D7EA7FE34A9415B4
  balance: 1568
  staked: 4704
  power: 1000000
\end{ttlst}\end{shellbox}
%
We already know the first one, that is, the gamete\index{gamete}.
Its private key is not stored in the docker container
but must be available to the person who started the container.
Normally, it is the key that was created before starting the node (with \texttt{moka keys create}
or similar) and
that is later bound to the storage address of the gamete (with \texttt{moka keys bind}). If you
followed the instructions in the previous sections, you should have a
\texttt{gamete.pem} file in your file system, for the gamete. With that pem file,
you have \emph{superuser} rights,
in the sense that you can, for instance,
open and close the faucet\index{faucet} (but only if you configured the new blockchain
with the \texttt{ALLOWS\_UNSIGNED\_FAUCET} option set to true).
Moreover, the gamete owns all cryptocurrency initially minted for the node (if any).
With that, you can create and fund as many new accounts
as you want and in general run any transaction you like.

There is a second account that has been created. Namely, the \emph{validator}\index{validator} account.
This is an externally owned account that gets remunerated for every
non-\texttt{@View}\index{View@{\texttt{View}}}
transaction run in the node and included in blockchain. At
the beginning, the balance of a validator is 0. This increases for each non-\texttt{@View}
transaction executed in the blockchain. Namely, the gas
consumed for such transactions gets forwarded to the validators of the blockchain, at its current price,
in proportion to their validation power\index{validation power}.
If there is only a single validator, everything goes to it.
It is important to note that only a portion of this prize lands, immediately, in full,
in the balance of the validators: the rest is \emph{staked}\index{staking}
for each validator, that is, kept in the validators contract
as a motivation for the validators to behave correctly. In the future, if a validator misbehaves
(that is, if it does not validate the transactions correctly or does not validate them at all) then
its stake will be reduced by a percent that is called \emph{slashing}\index{slashing}.
This is by default $1\%$ for validators that do not validate correctly and $0.5\%$ for
validators that do not validate at all (for instance, they are down).
The staked amount will be forwarded to a validator only when it will sell all its
validation power to another validator and stop being a validator.

The power of a validator expresses also the weight of its votes: in order to validate a transaction,
at least two thirds of the validation power must agree on the outcome of the transaction, or
otherwise the network will hang.
This mechanism is inherited from the underlying Tendermint engine and might be different
in other Hotmoka nodes in the future. However, the idea that power represents the weight
of a validator will likely remain.

Validators also receive a prize at the end of the creation of each block of the blockchain,
when coins are minted and distributed to the validators in proportion to their validation power.

Therefore, validator accounts receives payments for the validation of transactions
and the creation of new blocks.
But who controls this validators? It turns out that the \texttt{config-init} command
has created the key pair of the validator inside the configuration directory of the node.
You can see it if you access the \texttt{hotmoka\_tendermint} volume where the container operates.
You must be root to do that:
%
\begin{shellcommandbox}\begin{ttlst}
sudo ls /var/lib/docker/volumes/hotmoka_tendermint/_data
\end{ttlst}\end{shellcommandbox}
%
\begin{shellbox}\begin{ttlst}
consensus_config.toml  local_config.toml  tendermint_config  validator.pem
\end{ttlst}\end{shellbox}
%
In alternative, you can use the \texttt{docker exec} command to run a command inside the container.
You do not need to be root, but need to know the id of the running container (\texttt{docker ps} might help you):
%
\begin{shellcommandbox}\begin{ttlst}
docker exec c1407e499ad67465318704da1fcb6e9b88ee94faceb7c4b86e00ab4775590b3f /bin/ls hotmoka_tendermint
\end{ttlst}\end{shellcommandbox}
%
\begin{shellbox}\begin{ttlst}
consensus_config.toml
local_config.toml
tendermint_config
validator.pem
\end{ttlst}\end{shellbox}
%
Who owns that key controls the validator. Therefore, you might wanto to move it in your host machine:
%
\begin{shellcommandbox}\begin{ttlst}
docker cp c1407e499:/home/hotmoka/hotmoka_tendermint/validator.pem .
\end{ttlst}\end{shellcommandbox}
%
\begin{shellbox}\begin{ttlst}
Successfully copied 2.05kB
\end{ttlst}\end{shellbox}

As a final remark about the key of the validator, note that it \emph{must} be the same
key that the underlying Tendermint engine uses in order to identify the node in the network
and vote for validation. If that is not the case, the validator account in the
manifest will not be recognized as a working validator and will be slashed for
not behaving. Eventually, it will be expulsed from the set of validators.
Tendermint stores the key that it uses to identify the node in another file, inside
its configuration, and in JSON format:
%
\begin{shellcommandbox}\begin{ttlst}
docker exec c1407e499 /bin/ls hotmoka_tendermint/tendermint_config/config
\end{ttlst}\end{shellcommandbox}
%
\begin{shellbox}\begin{ttlst}
config.toml
genesis.json
node_key.json
priv_validator_key.json
\end{ttlst}\end{shellbox}
%
This file must remain in the node, or otherwise Tendermint cannot vote for validation.
The docker \texttt{config-clone} and \texttt{config-new} scripts magically ensure that,
correctly, this file contains the same key
as \texttt{validator.pem}, although in a different format.

\subsection{How to become a validator}\label{subsec:become_validator}

The configuration of a Hotmoka Tendermint node, created thorugh the \texttt{config-clone}
or \texttt{config-new} commands,
contains a \texttt{validator.pem} key pair file, as said above. You can use that file to create an actual
account object, a candidate for becoming a validator of the network. You can do this with the
\texttt{moka nodes tendermint validators create}\index{moka@{\texttt{moka}}!nodes tendermint validators create@{\texttt{nodes tendermint validators create}}}
command, which is similar to
\texttt{moka accounts create} but creates instances of
\texttt{TendermintED25519Validator}\index{TendermintED25519Validator@{\texttt{TendermintED25519Validator}}}.
Once you have a validator object for your \texttt{validator.pem}, you can become an actual validator
when one of the existing validators decides to sell (part of) its shares and you buy it.
The seller validator creates a sale offer with the
\texttt{moka nodes tendermint validators sell}\index{moka@{\texttt{moka}}!nodes tendermint validators sell@{\texttt{nodes tendermint validators sell}}}
command. The sale offer will be visible in the manifest of the node and can be bought with
the \texttt{moka nodes tendermint validators buy}\index{moka@{\texttt{moka}}!nodes tendermint validators buy@{\texttt{nodes tendermint validators buy}}}
command. After that, the buyer will be a new validator of the network.

\begin{commentbox}
It is important to note that a validator node is assumed to be reachable from the outside world
and up, also overnight. Most home desktop computers, connected through a modem, have no public IP
that can be used to reach them from the outside world (the public IP of the modem reaches the modem
itself, not the computer). Therefore, if you try to use a home machine to become a validator, your
machine will be immediately slashed for being uneachable and it will be removed from the set of validators,
immediately after becoming one. You need a machine with a public IP for being a validator, such as
for instance a machine on a rental service.
\end{commentbox}